Global multivariate model learning from hierarchically correlated data

### 2021-01-25 main.tex
Intro draft

\usepackage{graphicx}% Include figure files
\usepackage{dcolumn}% Align table columns on decimal point
\usepackage{bm}% bold math
\usepackage{color}
\usepackage{epstopdf} 
\usepackage{amsmath,amsfonts,amssymb,amsthm}
% \usepackage{hyperref}
\usepackage{algorithm}
\usepackage{algorithmic}
\usepackage{braket}
\usepackage[sans]{dsfont}
\usepackage{mathbbol}
\usepackage{bbm}

@@

% Pierre
\newcommand{\pierre}[1]{{\color{red}Pierre: #1}}
\newcommand{\curlynormalpar}[1]{\exp\left\{-\frac{1}{2}\left( #1 \right)\right\}}
\newcommand{\curlynormal}[1]{\exp\left\{-\frac{1}{2} #1 \right\}}
\newcommand{\iSig}{\bm{\Sigma^{-1}}}
\newcommand{\iC}{\bm{C}^{-1}}
\newcommand{\av}[1]{\left\langle #1 \right\rangle}
\newcommand{\vsa}{\vec{s}_a}
\newcommand{\vuka}{\vec{u}_{ka}}

@@

\title{Inference of global coevolutionary models from hierarchical correlated data.}
\date{}

\author{Edwin Rodriguez Horta} 
\affiliation{Group of Complex Systems and Statistical Physics, Department of Theoretical Physics, University of Havana, Cuba}
\affiliation{Laboratoire de Biologie Computationnelle et Quantitative (LCQB), Sorbonne Université, France}
%
\author{Pierre Barrat-Charlaix} 
\affiliation{Biozentrum, Universität Basel, Switzerland}
%
\author{Alejandro Lage} 
\affiliation{Group of Complex Systems and Statistical Physics, Department of Theoretical Physics, University of Havana, Cuba}
%
\author{Martin Weigt} 
\email{Correspondence to: Martin Weigt, \bf{martin.weigt@upmc.fr}}
\affiliation{Laboratoire de Biologie Computationnelle et Quantitative, Sorbonne Université, France}

@@

\section{Introduction}
\label{sec:int}
Theory and applications of the inverse problem of statistical physics has had great development in recent years since the availability of large amounts of data on a microscopic scale produced by the continuous technological progress. In this context statistical physics paths: from model parameters to  observables, is reversed, toward find global models which describe the statistics of  a set of observed  configurations \cite{Inverse_problem_Berg}.

@@

@@

We consider a system characterized by $L$ continuous degrees of freedom and whose state is fully described by a vector $\vx$ of length $L$. 
These degrees of freedom can be continuous phenotypic traits of some living organism, or the sequence of a gene or a protein if a continuous approximation is made. 
At equilibrium, $\vx$ is normally distributed:
\begin{equation}
	P_{eq}(\vx) = \frac{1}{Z(\bm{J})}\curlynormal{\vx^T\bm{J}\vx},
	\label{eq:eq_distribution}
\end{equation}
where $\bm{J}$ is the symmetric positive definite \emph{coupling matrix} and $Z(\bm{J})$ is a normalization constant. 
We are interested in inferring this coupling matrix from a given amount of observed states $\vx$ of the system.
If these observations were independent from each other, $\bm{J}$ would simply be equal to the inverse of the \emph{covariance matrix} of the data, written $\bm{C}$.

However, we are concerned here with the case where these observations are not independent. 
On the contrary, they are the result of a dynamical process which took place during a finite amount of time, and is thus not necessarily equilibrated. 
This dynamical process is described below.

We suppose that the considered system evolves according to the following Langevin equation 
\begin{equation}
	\gamma^{-1}\frac{\text{d}\vx}{\text{d}t} = - \bm{J}\vx + \vec{\xi}(t).
	\label{eq:langevin}
\end{equation}
Here, $\vec{\xi}(t)$ is a vector of uncorrelated white noises, and $\gamma^{-1}$ is the characteristic timescale governing the dynamics. 
In short, equation~\ref{eq:langevin} states that the system described by $\vx$ undergoes Brownian motion in an energy landscape characterized by the coupling matrix $\bm{J}$.

We are not interested in $\vx$ directly, but rather by its probability distribution $P(\vx,t\vert\,\vx_0)$, that is the probability to find the system in state $\vx$ knowing it was in state $\vx_0$ and evolved for time $t$. 
The Fokker-Planck equation corresponding to equation~\ref{eq:langevin} is straightforward to write: 
\begin{equation}
	\gamma \partial_t P = \left(-\sum_{a,b=1}^L\frac{\partial}{\partial x_a} J_{ab}x_b + \sum_{a=1}^L\frac{\partial^2}{\partial x_a^2}\right) P, 
	\label{eq:FP}
\end{equation}
where the parenthesized expression on the right hand side is understood as an operator on $P$. 
The solution to~\ref{eq:FP} is a multivariate normal distribution~\cite{singh2017multiOU}:
\begin{equation}
	P(\vx, \Delta t|\, \vx_0) = \left((2\pi)^N\det\iSig\right)^{-1/2}
	\curlynormal{(\vx-\vec{\mu})^T\iSig(\vx- \vec{\mu})},
	\label{eq:OUpropagator}
\end{equation}
where we introduce matrices $\Sig$ and $\Lam$ as well as vector $\vec{\mu}$: 
\begin{equation}
	\Lam = e^{-\gamma\bm{J}}, \qquad \vec{\mu} = \Lam^{\Delta t} \vx_0, \qquad \Sig = \bm{J}^{-1}(\mathbb{1} - \Lam^{2\Delta t}).
	\label{eq:def_lambda}
\end{equation}
Equations~\ref{eq:OUpropagator} and~\ref{eq:def_lambda} define a multivariate \emph{Ornstein-Uhlenbeck} process (OU).

It is important to note that since matrix $\Lam$  is the exponential of $-\bm{J}$, it is symmetric, has strictly negative eigenvalues and commutes with $\bm{J}$. 
We also underline the fact that $\Sig$ and $\vec{mu}$ depend on $\Delta t$, although this dependence is absent in our notation.

By taking $\gamma\Delta t \gg 1$ and using the fact that $\bm{J}$ has strictly positive eigenvalues , one immediately recovers equation~\ref{eq:eq_distribution}. 
This allows us to compute the joint distribution of two configurations $\vx_1$ and $\vx_2$ separated by a time $\Delta t$ by multiplying equations~\ref{eq:eq_distribution} and~\ref{eq:OUpropagator}: 
\begin{equation}
	P(\vx_1, \vx_2\vert\,\Delta t) \propto \curlynormalpar{\vx_1^T\iSig\vx_1 + \vx_2^T\iSig\vx_2 - 2\vx_1^T\Lam^{\Delta t}\iSig\vx_2}. 
	\label{eq:OUjoint}	
\end{equation}
This last equation is important as it allows one to compute the joint covariance of configurations $\vx_1$ and $\vx_2$. 
The probability distribution in~\ref{eq:OUjoint} is a normal distribution whose inverse covariance matrix is defined by blocks: $\Sig$ on the diagonal and $-\Lam^{\Delta t}\Sig$ off-diagonal. 
By inverting this block matrix, given that $\Lam$ and $\Sig$ commute and are invertible, one obtains the following covariance: 
\begin{equation}
	\langle \vx_1\vx_2^T \rangle_{\Delta t} = \Lam^{\Delta t}\bm{J}^{-1} = \Lam^{\Delta t}\bm{C}. 
	\label{eq:pairwisecov}
\end{equation}
As said above, the aim of this work is to infer matrix $\bm{J}$ from a set of observations $\vx$ that are generated this OU process. 
More precisely, any two observations $\vx_1$ and $\vx_2$ will be separated by a time $\Delta t$ during which the OU process took place. 
Equation~\ref{eq:pairwisecov} then allows us to distinguish two regimes. 
Let us call $\{\rho_a\}$ the eigenvalues of $\bm{J}$. 
Using the definition of $\Lam$, we immediatly have that the eigenvalues of matrix $\Lam^{\Delta t}\bm{C}$, \ie  the left hand side of equation~\ref{eq:pairwisecov}, are of the form $\rho_a^{-1}e^{-\gamma\rho_a\Delta t}$.

Since the $\rho_a$ are positive, the eigenvalues of $\Lam^{\Delta t}\bm{C}$ vanish exponentially for large enough times. 
The timescale of this exponential decrease is set by the characteristic time $\tau_c^{-1} = \gamma\rho_{min}$, where $\rho_{min}$ is the smallest eigenvalue of $\bm{J}$. 
Thus, for $\Delta t / \tau_c \gg 1$, $\vx_1$ and $\vx_2$ are uncorrelated. 
If this is verified for all pairs of observations, the regime is that of \emph{weakly correlated} data, and the inference of $\bm{J}$ can simply be performed by inverting the empirical correlation matrix. 
Inversely, for $\Delta t / \tau_c \ll 1$, $\vx_1$ and $\vx_2$ are highly correlated, defining a \emph{strongly correlated} regime. 
It should be noted that when $\Delta t = 0$, the joint correlation matrix of $\vx_1$ and $\vx_2$ becomes non invertible, and equation~\ref{eq:pairwisecov} becomes irrelevant.

\section{Statement of the problem}
\label{statement of the problem}

\begin{figure}[!htb]
	\includegraphics[width=0.6\textwidth]{Figures/tree.pdf}
	\caption{ \pierre{We should get a better figure here.}}
	\label{fig:sample_tree}
\end{figure}

The central problem discussed here is the inference of the probability distribution describing samples that are hierarchically correlated by a tree. 
Formally, we assume that the data consists of $N$ real valued vectors of length $L$, noted $\{\vx_i\}$ with $i\in\{1\ldots N\}$. 
Taken individually, the $\vx_i$s are normally distributed according to equation~\ref{eq:eq_distribution}. with mean $0$.
By construction, the equilibrium correlation between components of vectors $\vx$ is given by the inverse of the couplings, that is $\langle x^a x^b\rangle-\langle x^a\rangle\langle x^b\rangle = \bm C_{ab} = (\bm J^{-1})_{ab}$. 
This implies that inferring the coupling matrix defining the probability distribution amounts to finding the \emph{equilibrium} covariance matrix $\mathbf{C}$.

However, this covariance cannot be directly measured as we consider observations that are not independently distributed.
Instead, the set of measured configurations $\{\vx_i\}$ is the result of an Ornstein-Uhlenbeck (OU) process taking place on a tree $\mathcal{T}$, such as the one shown in figure~\ref{fig:sample_tree}. 
The configuration $\vx_i$ of each node $c$ in the tree is the result of the evolution given by the equation~\ref{eq:langevin}, with initial condition $\vx_{a(i)}$, $a(i)$ being the ancestral node of $i$, and running time $\Delta t_{ia(i)}$, that is the \emph{branch length} between $a(i)$ and $i$. 
This process is initialized with a root note $\vx_r$ that is distributed according to equation~\ref{eq:eq_distribution}.
Observed data points are leaves of this tree, which is assumed to be perfectly known.
The nature of the chosen evolution process implies that the distribution of a leaf configuration $\vx_i$ conditioned on the knowledge of another leaf configuration $\vx_j$ is given by equation~\ref{eq:OUpropagator}, where $\Delta t=\Delta t_{ij}$ represents the total branch length separating leaf $i$ and leaf $j$.

The OU process is fully characterized by the quadratic potential $\bm J = \bm C^{-1}$ and a rate $\gamma$. 
Hence, the configurations of the leaves, that is the data, are determined only by $\bm C$, $\gamma$, and the tree $\mathcal{T}$. 
The aim of this work is to design a method to infer the most likely values of $\bm C$ and $\gamma$ from the knowledge of the data $\mathcal{D}=\{\vx_i\}$ and the shape of $\mathcal{T}$. 
There are two notable extreme cases to this problem. 
The first one is the $\Delta t_r/\tau_c\longrightarrow 0$ case, where $\Delta t_r$ is the typical branch length distance from leaves to root. 
Leaf configurations are then almost identical to the root, \emph{i.e.} $\vx_i\simeq\vx_r$, and inference of $\bm C$ becomes impossible. 
The second one is the $\Delta t_r/\tau_c\longrightarrow\infty$ case, where equilibration of the OU process takes place almost instantly when compared to branch lengths of $\mathcal{T}$.
Leaf configurations are then independently drawn from the equilibrium distribution $P_{eq}$, and $\bm C$ can be readily estimated from the empirical covariances. 
Here, we deal with the intermediary cases where the tree makes it impossible to simply estimate $\bm C$ from empirical covariances, but where $\tau_c$ is not large enough to remove all signatures of the energy potential.

We adopt a Bayesian approach to this inference problem by writing the probability of a given set of parameters as a function of the data using the following relation
\begin{equation}
	P(\bm C, \gamma\vert\mathcal{D},\mathcal{T}) \propto P(\mathcal{D}\vert\bm C, \gamma,\mathcal{T})\cdot P(\bm C, \gamma),
	\label{eq:bayesian_likelihood}
\end{equation}
where the proportionality constant does not depend on the parameters $\bm C$ or $\gamma$, and where $P(\bm C, \gamma)$ can be any arbitrarily chosen prior distribution. The difficulty in equation~\ref{eq:bayesian_likelihood} lies in the estimation of the term $P(\mathcal{D}\vert\bm C, \gamma, \mathcal{T})$, that is the likelihood of the data given a set of parameters $\bm C$ and $\gamma$ and the tree $\mathcal{T}$.  We detail the computation of this likelihood in the following section.

@@

The joint distribution of two configurations $\vx_1$ and $\vx_2$ separated by time $\Delta t$ is given by equation~\ref{eq:OUjoint} and corresponds to a joint normal distribution. 
This means that the vector $\vec{X}=[\vx_1, \vx_2]$, \emph{i.e.} the concatenation of vectors $\vx_1$ and $\vx_2$, follows a normal distribution with null means.
Of importance here is that this property of the OU process can be extended to the joint distribution of any subset nodes in a tree. 
In other words, if we now define $\vec{X}=[\vx_1,\ldots,\vx_N]$ to be the concatenation of the configurations of all leaves of $\mathcal{T}$, we can write the distribution of $\vec{X}$ as
\begin{equation}
	P(\vec{X}\vert\,\bm{C},\gamma,\mathcal{T}) = \left((2\pi)^{LN}\det\mathbb{G}\right)^{-\frac{1}{2}}\curlynormal{\vec{X}^T\mathbb{G}^{-1}\vec{X}},
	\label{eq:multiOU_jointall}
\end{equation}
where $\mathbb{G}$ is the \emph{joint correlation matrix} and depends on the tree as well as on $\bm{C}$ and $\gamma$.
This property is shown in more detail in section \ref{sec:joint_probability_of_leaves_is_a_gaussian} of the Supplementary Material.\\
Equation~\ref{eq:multiOU_jointall} allows us to compute the log-likelihood of the data $\vec{X}$ as a function of $\vec{X}$ itself and of the joint correlation matrix. 
Indeed, taking its logarithm immediatly gives
\begin{equation}
	\mathcal{L}_{\mathcal{D}}(\mathbb{G}) = -\frac 1 2 \log \det  \mathbb{G}  -\frac 1 2 \vec{X}^T\mathbb{G}^{-1} \vec{X} + \text{cte}. 
	\label{eq:likelihood_2}
\end{equation}

>>>>

\usepackage{graphicx}% Include figure files
\usepackage{dcolumn}% Align table columns on decimal point
\usepackage{bm}% bold math
\usepackage{color}
\usepackage{epstopdf} 
\usepackage{amsmath,amsfonts,amssymb,amsthm}
% \usepackage{hyperref}
\usepackage{algorithm}
\usepackage{algorithmic}
\usepackage{braket}
\usepackage[sans]{dsfont}
\usepackage{mathbbol}
\usepackage{bbm}
\usepackage{subfigure}

@@

% Pierre
\newcommand{\curlynormalpar}[1]{\exp\left\{-\frac{1}{2}\left( #1 \right)\right\}}
\newcommand{\curlynormal}[1]{\exp\left\{-\frac{1}{2} #1 \right\}}
\newcommand{\iSig}{\bm{\Sigma^{-1}}}
\newcommand{\iC}{\bm{C}^{-1}}
\newcommand{\av}[1]{\left\langle #1 \right\rangle}
\newcommand{\vsa}{\vec{s}_a}
\newcommand{\vuka}{\vec{u}_{ka}}

\newcommand{\sref}[1]{S\ref{#1}}
\newcommand{\pierre}[1]{{\color{red}Pierre: #1}}
\newcommand{\martin}[1]{{\color{blue}Martin: #1}}

@@

\title{Global multivariate model learning from hierarchically correlated data}
\date{}

\author{Edwin Rodriguez Horta} 
\affiliation{Sorbonne Universit\'{e}, CNRS, Institut de Biologie
  Paris-Seine, Laboratoire de Biologie Computationnelle et
  Quantitative -- LCQB, Paris, France}
\affiliation{Group of Complex Systems and Statistical Physics, Department of Theoretical Physics, University of Havana, Havana, Cuba}
%
\author{Pierre Barrat-Charlaix} 
\affiliation{Biozentrum, Universität Basel, Basel, Switzerland}
%
\author{Alejandro Lage} 
\affiliation{Group of Complex Systems and Statistical Physics, Department of Theoretical Physics, University of Havana, Havana, Cuba}
%
\author{Martin Weigt} 
\email{Correspondence to: Martin Weigt, \bf{martin.weigt@upmc.fr}}
\affiliation{Sorbonne Universit\'{e}, CNRS, Institut de Biologie
  Paris-Seine, Laboratoire de Biologie Computationnelle et
  Quantitative -- LCQB, Paris, France}

@@

\section{Introduction}
\label{sec:int}

In the last decades, biology has seen impressive progress in experimental techniques which has resulted in a large increase of available data. 
This is especially visible in the case of biological sequences,  with databases now harboring a vast amount of high-quality DNA or protein sequences. [UNIPROT/GENBANK]
This has in turn fuelled a development of quantitative methods to model forces guiding the evolution of sequences or other biological traits [REFS?]. 
A common idea in this context is that it is possible to use characteristics of homologous genes or organisms to construct quantitative models of the selection acting on them. 
A successful example is the representation of protein sequences by probabilistic models in the so-called DCA method [REF].

Models built in this way usually assume that biological characteristics of related sequences or organisms are evolving under the same constraints but are independent from each other in the statistical sense. 
This allows simple formulation for model learning. 
However, phylogenetic relations between evolving entities imply that these cannot be considered independent from each other, and doing so can result in important biases when learning models. 
For instance, it has been shown that phylogenetic relations between protein sequences induce non-trivial correlations that are not related to protein function [COLWELL paper].

In order to correct for these biases, it is necessary to disentangle the statistical effects due to selection forces and those due to phylogeny. 
However, this becomes a complex problem when dealing with complex models where traits under study are not independent from each other. 
Accounting for these biases then requires one to understand how correlated traits evolve along a tree. 
History of modeling this with BM or OU processes [Felsenstein, Hansen]. Also, field of PCM

Increasing data --> new quantitative techniques in biology. 
One big recent success: pairwise sequence models, blabla
However, models learned from data with phylogenetic correlations can suffer from biases. To account for these biases in the case of complex models, it's necessary to understand how correlated traits evolve along a tree (or: quantities that are modeled and typically not independent evolve along a tree)

There is a history of looking at multi parameter complex models and their relation with phylogeny. 
in this context, it is often assumed that quantitative traits that evolve under selection can be modeled using an OU process. 
This is usually done in the field of PCMs, with most common models being Brownian motion or OU processes.

Here, we want to understand how to find parameters values of an OU gaussian process from data correlated through a tree. 
This could be applied to the evolution of any correlated biological traits, but also potentially to protein sequences.

The MS is divided as follows: >>>

\paragraph{Old} % (fold)
\label{par:old}

Theory and applications of the inverse problem of statistical physics has had great development in recent years since the availability of large amounts of data on a microscopic scale produced by the continuous technological progress. In this context statistical physics paths: from model parameters to  observables, is reversed, toward find global models which describe the statistics of  a set of observed  configurations \cite{Inverse_problem_Berg}.

@@

% paragraph old (end)

@@

We consider a system characterized by $L$ continuous degrees of freedom and whose state is fully described by an $L$-dimensional vector $\vx\in\mathbb{R}^L$. These degrees of freedom can be continuous phenotypic traits of some living organism, or the sequence of a gene or a protein if a continuous approximation is made. 
At equilibrium, $\vx$ is assumed to be normally distributed,
\begin{equation}
	P_{eq}(\vx) = \frac{1}{Z(\bm{J})}\curlynormal{\vx^T\bm{J}\vx}\ ,
	\label{eq:eq_distribution}
\end{equation}
where $\bm{J}$ is the symmetric, positive definite \emph{coupling matrix} and $Z(\bm{J}) = \sqrt{(2\pi)^L/\det \bm{J}}$ is the normalization constant; the means of all components of $\vx$ are set to zero without loss of generality. 
We are interested in inferring the coupling matrix from a given amount of observed states $\vx$ of the system.
If these observations were independent from each other, due to the simple Gaussian form of Eq.~(\ref{eq:eq_distribution}), $\bm{J}$ would simply be equal to the inverse of the empirical \emph{covariance matrix} of the data, written $\bm{C}=\bm{J}^{-1}$.

However, here we are concerned with the case where observations are not independent. 
On the contrary, they result from a dynamical process taking place during a finite amount of time. 
Different datapoints are therefore correlated even if individually assumed to be at equilibrium. 
This dynamical process is described below.

We suppose that the considered system evolves according to the following Langevin equation 
\begin{equation}
	\gamma^{-1}\frac{\text{d}\vx}{\text{d}t} = - \bm{J}\vx + \vec{\xi}(t).
	\label{eq:langevin}
\end{equation}
Here, $\vec{\xi}(t)$ is a vector of uncorrelated white noise, and $\gamma^{-1}$ is the characteristic timescale governing the dynamics. 
In short, Eq.~(\ref{eq:langevin}) states that the system described by $\vx$ undergoes Brownian motion in a quadratic energy landscape characterized by the coupling matrix $\bm{J}$.

We are not interested in $\vx$ directly, but rather in its probability distribution $P(\vx\vert\,\vx_0,\Delta t)$, \ie~in the probability to find the system in state $\vx$ knowing it was in state $\vx_0$ some time $\Delta t$ before. 
The Fokker-Planck equation corresponding to Eq.~(\ref{eq:langevin}) is straightforward to write, 
\begin{equation}
	\gamma^{-1} \partial_t P = \left(-\sum_{a,b=1}^L\frac{\partial}{\partial x_a} J_{ab}x_b + \sum_{a=1}^L\frac{\partial^2}{\partial x_a^2}\right) P, 
	\label{eq:FP}
\end{equation}
where the parenthesized expression on the right hand side is understood as an operator acting on $P$. 
The solution to Eq.~(\ref{eq:FP}) is a multivariate normal distribution~\cite{singh2017multiOU}:
\begin{equation}
	P(\vx|\, \vx_0, \Delta t) = \left[(2\pi)^N\det\Sig\right]^{-1/2}
	\curlynormal{(\vx-\vec{\mu})^T\iSig(\vx- \vec{\mu})},
	\label{eq:OUpropagator}
\end{equation}
where we introduce the matrices $\Sig$ and $\Lam$ as well as the vector $\vec{\mu}$ as 
\begin{equation}
	\Lam = e^{-\gamma\bm{J}}, \qquad \vec{\mu} = \Lam^{\Delta t} \vx_0, \qquad \Sig = \bm{J}^{-1}(\mathbb{1} - \Lam^{2 \Delta t}).
	\label{eq:def_lambda}
\end{equation}
Eqs.~(\ref{eq:OUpropagator}) and~(\ref{eq:def_lambda}) define a multivariate \emph{Ornstein-Uhlenbeck} (OU) process.

Note that since matrix $\Lam$ is an exponential of $\bm{J}$, it is symmetric, has strictly positive eigenvalues and commutes with $\bm{J}$. 
We also underline that $\Sig$ and $\vec{\mu}$ depend on $\Delta t$, although this dependence is not explicitly written in our notation to make it less heavy.

By taking $\gamma\Delta t \gg 1$ and using the fact that $\bm{J}$ has strictly positive eigenvalues, one immediately recovers Eq.~(\ref{eq:eq_distribution}). 
We can thus compute the joint distribution of two configurations $\vx_1$ and $\vx_2$ separated by time difference $\Delta t$ by multiplying Eqs.~(\ref{eq:eq_distribution}) and~(\ref{eq:OUpropagator}), 
\begin{eqnarray}
    P(\vx_1, \vx_2\vert\,\Delta t) &= & P(\vx_1 |\, \vx_2,\Delta t) \times P_{eq}(\vx_2)
    \nonumber\\
    &\propto& \curlynormalpar{\vx_1^T\iSig\vx_1 + \vx_2^T\iSig\vx_2 - 2\vx_1^T\Lam^{\Delta t}\iSig\vx_2}. 
	\label{eq:OUjoint}	
\end{eqnarray}
This equation illustrates the \emph{time reversibility} of the OU process. 
Indeed, the distribution does not depend on which configuration $\vx_1$ or $\vx_2$ came first.

Equation \eqref{eq:OUjoint} allows for computing the joint covariance of the correlated equilibrium configurations $\vx_1$ and $\vx_2$. 
The probability distribution in Eq.~(\ref{eq:OUjoint}) is normal with an inverse covariance matrix defined by blocks: $\Sig$ on the diagonal and $-\Lam^{\Delta t}\Sig$ off-diagonal. 
By inverting this block matrix, given that $\Lam$ and $\Sig$ commute and are invertible, one obtains the following covariance: 
\begin{equation}
	\langle \vx_1\vx_2^T \rangle_{\Delta t} = \Lam^{\Delta t}\bm{J}^{-1} = \Lam^{\Delta t}\bm{C}. 
	\label{eq:pairwisecov}
\end{equation}

Eq.~(\ref{eq:pairwisecov}) allows us to readily distinguish two regimes. 
Let us call $\rho_a$ the eigenvalues of $\bm{J}$. 
The eigenvalues of $\Lam^{\Delta t}\bm{C}$ are then equal to $\rho_a^{-1}e^{-\gamma\rho_a\Delta t}$. 
Since all $\rho_a$ are positive, the eigenvalues of $\Lam^{\Delta t}\bm{C}$ vanish exponentially over time. The slowest timescale of exponential decay is set by $\tau_c^{-1} = \gamma\rho_{min}$, with $\rho_{min}$ being the smallest eigenvalue of $\bm{J}$. 
Thus, for $\Delta t / \tau_c \gg 1$, $\vx_1$ and $\vx_2$ are uncorrelated. 
If this is verified for all pairs of observations $\vx_i$ and $\vx_j$, the regime is that of \emph{uncorrelated} data -- the inference of $\bm{J}$ can simply be performed by inverting the empirical covariance matrix extracted from the data. 
Inversely, for $\Delta t / \tau_c \ll 1$, $\vx_1$ and $\vx_2$ are highly correlated, defining a \emph{strongly correlated} regime. 
It should be noted that for $\Delta t = 0$, the joint correlation matrix of $\vx_1$ and $\vx_2$ becomes non invertible, and Eq.~(\ref{eq:pairwisecov}) becomes irrelevant. Actually, $\vx_1$ and $\vx_2$ coincide at that point, \ie~we have $P(\vx_1,\vx_2|\,\Delta t=0) = P_{eq}(\vx_1) \times \delta(\vx_1 - \vx_2)$ using the $L$-dimensional Dirac distribution.

\section{Methods}
\label{sec:methods}

\subsection{Statement of the problem}
\label{sub:statement_of_the_problem}

\begin{figure}[!htb]
	\includegraphics[width=0.8\textwidth]{Figures/tree.pdf}
	\caption{Schematic representation of a tree $\mathcal{T}$ underlying the data generating process. The process starts at the root node $r$ with a configuration $\vx_r$ sampled from $P_{eq}(\vx_r)$. The dynamics consist in independent realizations of the OU process on all branches from ancestral nodes $a(n)$ to child nodes $n$ over times corresponding to the branch length $\Delta t_{n,a(n)}$, initialized in the ancestral configuration $\vx_{a(n)}$. The observable data only consist of configurations of the leaf nodes (grey circles in the figure), while configurations of ancestral nodes remain unknown. There are no restrictions on the topology of tree $\mathcal{T}$ and the length of the branches.}
	\label{fig:sample_tree}
\end{figure}

The problem discussed here is the inference of the probability distribution describing samples that are hierarchically correlated by a tree, cf.~Fig.~\ref{fig:sample_tree}. 
Formally, we assume that the data consists of $N$ real-valued vectors of length $L$, denoted $\{\vx_i\}\in\mathbb{R}^L$ with $i=1,...,N$. 
Taken individually, the $\vx_i$ are distributed according to Eq.~\ref{eq:eq_distribution}, \ie~according to a multivariate Gaussian of zero mean and covariance $\bm C$.
By construction, the equilibrium covariance between any pair of elements of a given vector $\vx=(x^1,...,x^L)$ is given by the inverse of the coupling matrix: $\langle x^a x^b\rangle-\langle x^a\rangle\langle x^b\rangle = \bm C_{ab} = (\bm J^{-1})_{ab}$ for all $a,b=1,...,L$. 
This implies that inferring the coupling matrix defining the probability distribution amounts to finding the \emph{equilibrium} covariance matrix $\mathbf{C}$.

However, this covariance cannot be directly measured as we consider observations that are not independently distributed.
Instead, the set of measured configurations $\{\vx_i\}_{i=1,...,N}$ is the result of an Ornstein-Uhlenbeck (OU) process taking place on a tree $\mathcal{T}$, as is illustrated in Fig.~\ref{fig:sample_tree}:
\begin{itemize}
    \item The process starts at the root node $r$ with a state vector $\vx_r$ drawn from the equilibrium distribution $P_{eq}$.
    \item On each branch $(n,a(n))$ of length $\Delta t_{n,a(n)}$ connecting node $n$ with its ancestral node $a(n)$, the dynamics follow Eq.~(\ref{eq:langevin}), starting from initial condition $\vx_{a(n)}$, and running for time $\Delta t_{n,a(n)}$. In other words, given the state $\vx_{a(n)}$ of the ancestral node, $\vx_n$ is sampled from $P(\vx_n |\, \vx_{a(n)}, \Delta t_{n,a(n)})$, see Eq.~(\ref{eq:OUpropagator})
    \item As a consequence, OU processes on branches stemming from common ancestral node evolve independently, but from an identical initial condition.
    \item Observed data vectors correspond to the states of the leaves of the tree at the end of this process. The states of the internal nodes are not part of the observed data and remain unknown. 
\end{itemize}
This process is thought to represent the evolution of biologic traits along a phylogenetic tree, with the leave nodes corresponding to traits observed in today's species. Note that due to the reversible nature of our OU process, the joint probability of any pair of leaf configurations $\vx_i$ and $\vx_j$, with $i,j\in\{1,...,N\}$, is given by  $P(\vx_i,\vx_2|\,\Delta t_{ij})$ (Eq.~(\ref{eq:OUjoint}), with $\Delta t_{ij}$ denoting the total branch length of the path connecting $i$ and $j$ in the tree.

The OU process is characterized by the quadratic potential $\bm J = \bm C^{-1}$ and the rate $\gamma$. 
Hence, the joint statistics of the leaf configurations $\{\vx_i\}_{i=1,...,N}$ (\ie~the data) is fully determined by $\bm C$, $\gamma$, and the tree $\mathcal{T}$. 
The aim of this work is to derive a method for inferring the most likely values of $\bm C$ and $\gamma$ given the knowledge of the data $\mathcal{D}=\{\vx_i\}_{i=1,...,N}$ and the underlying tree $\mathcal{T}$. We consider here that both the topology and the branch lengths of $\mathcal{T}$ are known.

This problem shows two notable extreme cases: 
The first one is the case where the typical branch length of the tree is short compared to the timescales of the OU process.
As a consequence, leaf configurations are close to identical to the root, \ie~$\vx_i\simeq\vx_r$, and the inference of $\bm C$ becomes impossible. 
The second one is the opposite case where the typical branch length of the tree is long compared to the longest timescale of the OU process $\tau_c$.
In this case, the configuration of a child node is close to independent from that of its ancestor, and leaf configurations can be considered as independent samples from the equilibrium distribution $P_{eq}$.
$\bm C$ can then be readily estimated by computing the empirical covariance matrix. 
We are interested here in the intermediate regime where substantial tree-mediated correlations between data
make it impossible to simply estimate $\bm C$ with the empirical covariance, but the depth of the tree introduces enough variability in the data for one to hope of reconstructing the energy potential $\bm{J}$.

We adopt a Bayesian inference approach by writing the probability of a given set of parameters $\{\bm C, \gamma\}$ given the data $\{\mathcal{D},\mathcal{T}\}$ using Bayes' equation
\begin{equation}
	P(\bm C, \gamma\vert\mathcal{D},\mathcal{T}) \propto P(\mathcal{D}\vert\bm C, \gamma,\mathcal{T})\cdot P(\bm C, \gamma),
	\label{eq:bayesian_likelihood}
\end{equation}
with the proportionality constant not depending on the parameters $\{\bm C, \gamma\}$. Here, $P(\bm C, \gamma)$ can be any arbitrarily chosen prior distribution. The difficulty in Eq.~(\ref{eq:bayesian_likelihood}) lies in the estimation of the likelihood $P(\mathcal{D}\vert\bm C, \gamma, \mathcal{T})$,
\ie~of the joint probability of the datapoints $\mathcal{D}=\{\vx_i\}_{i=1,...,N}$ for an OU process given by its parameters $\{\bm C, \gamma\}$ and the tree $\mathcal{T}$.  We detail the computation of this probability in the following section.

@@

The joint distribution of two configurations $\vx_1$ and $\vx_2$ separated by time $\Delta t$ is given by Eq.~(\ref{eq:OUjoint}) and corresponds to a joint normal distribution. 
This means that the vector $\vec{X}=[\vx_1, \vx_2]$, \emph{i.e.} the concatenation of vectors $\vx_1$ and $\vx_2$, follows a normal distribution with zero mean and variance described above in Eqs.~(\ref{eq:def_lambda}).
Of importance here is that this property of the OU process can be extended to the joint distribution of any subset of nodes in a tree. 
In other words, if we now define $\vec{X}=[\vx_1,\ldots,\vx_N]$ to be the concatenation of all configurations in our dataset $\mathcal D$, we can write the distribution of $\vec{X}$ as
\begin{equation}
	P(\vec{X}\vert\,\bm{C},\gamma,\mathcal{T}) = \left((2\pi)^{LN}\det\mathbb{G}\right)^{-\frac{1}{2}}\curlynormal{\vec{X}^T\mathbb{G}^{-1}\vec{X}},
	\label{eq:multiOU_jointall}
\end{equation}
where $\mathbb{G}$ is the \emph{joint covariance matrix} and depends on the tree as well as on $\bm{C}$ and $\gamma$.

The joint covariance matrix is a matrix of dimension $(L\cdot N)\times(L\cdot N)$, built by $N\times N$ blocks of size $L\times L$ with entries
\begin{equation}
	\mathbb{G}_{ij}(a,b) = \av{x_i^a x_j^b} - \av{x_i^a}\av{x_j^b},\ \ \ \ i,j\in\{1,...,N\}; a,b\in\{1,...,L\},
	\label{eq:jointcov_1}
\end{equation}
where the (zero) marginals $\av{x_i^a}$ and $\av{x_j^b}$ are explicitly written for clarity.
Each block $\mathbb{G}_{ij}$ is describing the connected correlations between two data vectors $\vx_i$ and $\vx_j$, which are separated by time $\Delta t_{ij}$, resulting as the sum of all branch lengths of the path connecting $i$ and $j$ on tree $\mathcal{T}$. Because the OU process is time reversible, we can directly apply Eq.~(\ref{eq:pairwisecov}) and give all blocks of $\mathbb{G}$ in closed form, 
\begin{equation}
	\mathbb{G}_{ij} =
	\begin{cases}
		\bm C & \text{if $i=j$}\\
		\Lam^{\Delta t_{ij}}{\bm C} & \text{otherwise}  ,
	\end{cases}
	\label{eq:jointcov_2}
\end{equation}
using the (currently unknown) covariance matrix $\bm C$ of a single equilibrium vector $\vx$. We remind here that $\Lam = e^{-\gamma\iC}$ depends only on $\gamma$ and $\bm C$, and commutes with $\bm C$. As a direct consequence, all blocks $\mathbb{G}_{ij}$ commute with each other and with $\bm C$.
